|

### 2024-12-17 Drafts/Paper_1/01-analysis_no_age_correction.Rmd
Working on Discussion update
All comments from the summer are now integrated

prop.slow.alaska <- round(100*n.alaska.slow/n.alaska,digits=0)
prop.slow.n24 <- round(100*n.nwa.slow/n.nafo.24,digits=0)

@@

r0.boxplt.6 <- ggplot(pop.dam.all.6) + geom_boxplot(aes(x=R0.opt,interaction(group,Stock.no.species),fill=group,color=group)) +
                                           geom_point(aes(x=med.r0,interaction(group,Stock.no.species)),color="firebrick2",fill="firebrick2",shape=8) +
                                           scale_y_discrete(NULL, guide = "axis_nested") +
                                           scale_color_manual(values = c("grey","blue")) +
                                           scale_fill_manual(values = c("grey",'blue')) +
                                       #scale_x_discrete(NULL, guide = "axis_nested") + 
                                        scale_x_log10(name = "Lifetime Reproductive Output")+
                                       geom_vline(xintercept = 1,linetype='dashed',color='blue')+

>>>>

prop.slow.alaska <- round(100*n.alaska.slow/n.alaska,digits=0)
prop.slow.n24 <- round(100*n.nwa.slow/n.nafo.24,digits=0)
# standard deviation in which rebuilding time was also above the low category
sd.rebuild.safe <- ceiling(max(lam.sd.p.rec$lam.no.fish.sd[lam.sd.p.rec$prop.cat == "Low"])*100)/100
lam.rebuild.safe <- ceiling(max(lam.sd.p.rec$lam.no.fish.med[lam.sd.p.rec$prop.cat == "Low"])*100)/100

@@

r0.boxplt.6 <- ggplot(pop.dam.all.6) + geom_boxplot(aes(x=R0.opt,interaction(group,Stock.no.species),fill=group,color=group)) +
                                           geom_point(aes(x=med.r0,interaction(group,Stock.no.species)),color="firebrick2",fill="firebrick2",shape=8) +
                                           scale_y_discrete(NULL, guide = "axis_nested") +
                                           scale_color_manual(values = c("grey","blue")) +
                                           scale_fill_manual(values = c("grey",'blue')) +
                                       #scale_x_discrete(NULL, guide = "axis_nested") + 
                                        scale_x_log10(name = "Lifetime Reproductive Output")+
                                       geom_vline(xintercept = 1,linetype='dashed',color='blue')

### 2024-12-17 Drafts/Paper_1/03-methods.Rmd
Working on Discussion update
All comments from the summer are now integrated

$$
1 = \sum_{a=\alpha}^{\omega} \lambda_y^{-a} s_{y,a} f_{y,a} dx
$$

Where $a$ is the age, $\alpha$ is the age at first reproduction, $\omega$ is the maximum age, $\lambda$ is the discrete rate of population growth ($\lambda = e^r$), $s_a$ is the survivorship to age $a$, and $f_a$ is the fecundity of individuals of age $a$. This equation is solved for each year ($y$) in which the data are available. Note that the discrete time Euler-Lotka equation can be shown to be the characteristic polynomial of the Leslie Matrix, where $f_0$ is the fecundity of the first age class with mature individuals [@leslieUseMatricesCertain1945, @caswellMatrixPopulationModels2001]. Survivorship-at-age and fecundity-at-age were calculated using the available stock assessment data, which enabled $\lambda$ to be estimated; note that this assumed that each stock represented a discrete population with individual dynamics.

@@

$$
RPA_{y,a} =  SSB_{(y-ra),a} RPS_{y}
$$
Fish are expected to have variability in reproductive success by age (ref, ref). However, we had to make the simplifying assumption that each kilogram of spawning biomass contributed the same to the total total reproductive output, irrespective of age. We calculated the age-specific fecundity ($f_{y,a}$) as the number of offspring produced by each spawner in the population. The number of spawners in year $y$ and age class $a$ ($NS_{y,a]$) is the proportion of mature individuals in each age class ($Mat_{y,a}$) multiplied by the number of individuals in each age class:

@@

<!-- This method does not account for any changes in reproductive success by age, each kilogram of spawning biomass contributes the same to the total total reproductive output, irrespective of their age. Finally, the age-specific fecundity ($f_{y,a}$) is calculated as number of offspring produced by each individual in the population.  -->

@@

$$
s_{y,a} =  s_{y,(a-1)} \times l_{y,(a-1)}
$$
The availability of catch-at-age time series in the data repository meant that fishing mortality and natural mortality rates were separable. Given this method provides separate estimates of $F$ and $M$ it was possible to calculate survival as well as theoretical population growth rates using the $M$ and fecundity ($\lambda_{F=0}$) estimates from the reconstruction. We recognize that the age structure in the underlying time series are still influenced by fishing pressure and that populations are at varying size. This means that ($\lambda_{F=0}$) does not represent the maximum intrinsic rate of abundance increase [e.g., r_max; CITE]. However, our theoretical annual growth rates in the absence of fishing represent the productivity that the population could have achieved in each specific year, if there had not been a fishery in that year.

@@

One limitation of calculating survivorship-to-age and fecundity-at-age using the archived data in the repository is that it is not straightforward to propagate observation error in each time series of data. To ensure that variability did not accumulate into unreasonable predictions for annual population productivity, we fit the Euler-Lotka model using an iterative minimization process, where predicted abundance in the following year:

@@

Life table analyses make no assumptions about population regulation and thus do not incorporate a density-dependent relationship between spawners and recruits [@mcallisterUsingDemographicMethods2001]. However, our estimated productivity parameters could be explored for evidence of density-dependence by modelling each stock’s annual population growth rate against relative population size (expressed as a proportion of maximum abundance) in the same year. Negative density dependence is observed when $\lambda$ increases as abundance declines [CITE] which is the type of compensatory response described by typical spawner-recruit relationships (ref, ref). Positive density dependence occurs when λ declines with increasing abundance, which is also called an Allee effect when it occurs at low abundance (ref, ref). Density independence occurs when λ shows no relationship with changes in abundance. We evaluated both the realized productivity estimates ($\lambda_{F>0}$) and the theoretical productivity values excluding the impact of fishing mortality ($\lambda_{F=0}$).

Life table analyses make no assumptions about population regulation and thus do not incorporate a density-dependent relationship between spawners and recruits (Reference?). However, the data could be explored for evidence of density-dependence by modelling each stock's estimated annual population growth rate against the relative population size (proportion of maximum abundance) that year. This was done twice, once using  ($\lambda_{F=0}$) and once using ($\lambda_{F>0}$).  Negative density dependence is observed when $\lambda$ increases as abundance declines, positive density dependence occurs when $\lambda$ declines with increasing abundance, while density independence is observed when $\lambda$ is not impacted by changes in abundance.

@@

The annual fecundity and natural mortality values varied considerably for each stock. To account for this variability when evaluating recovery potential, we used a simulated population projection. For `r n.dt.sims` simulations, we took a random sample of the annual values for fecundity and natural mortality, used the life table reconstruction to estimate $\lambda$, and calculated abundance in the following year. Abundance of each stock was projected over `r `n.sim.years` years, and the proportion of the simulations in which the population size had doubled was recorded each year. We considered stocks to have high recovery potential if 75% of simulated trajectories doubled, medium potential if 50-74% of simulations doubled, and low recovery potential if < 50% of simulated trajectories doubled, calculated at 10 and 20 years into the projection. For simplicity, we did not incorporate density-dependent effects (e.g., potential population decline as abundance exceeds carrying capacity) or any correlation (either temporally or between fecundity and natural mortality). Once a population had doubled, it was considered to have remained at that abundance or greater for the remainder of the projection. This means that the metric gave the earliest possible doubling time for each simulation, assuming no future fishing mortality.

>>>>

$$
1 = \sum_{a=\alpha}^{\omega} \lambda_y^{-a} s_{y,a} f_{y,a} dx
$$
Where $a$ is the age, $\alpha$ is the age at first reproduction, $\omega$ is the maximum age, $\lambda$ is the discrete rate of population growth ($\lambda = e^r$), $s_a$ is the survivorship to age $a$, and $f_a$ is the fecundity of individuals of age $a$. This equation is solved for each year ($y$) in which the data are available. Note that the discrete time Euler-Lotka equation can be shown to be the characteristic polynomial of the Leslie Matrix, where $f_0$ is the fecundity of the first age class with mature individuals [@leslieUseMatricesCertain1945, @caswellMatrixPopulationModels2001]. Survivorship-at-age and fecundity-at-age were calculated using the available stock assessment data, which enabled $\lambda$ to be estimated; note that this assumed that each stock represented a discrete population with individual dynamics.

@@

$$
RPA_{y,a} =  SSB_{(y-ra),a} RPS_{y}
$$
Fish are expected to have variability in reproductive success by age (ref, ref). However, we had to make the simplifying assumption that each kilogram of spawning biomass contributed the same to the total total reproductive output, irrespective of age. We calculated the age-specific fecundity ($f_{y,a}$) as the number of offspring produced by each spawner in the stock. The number of spawners in year $y$ and age class $a$ ($NS_{y,a]$) is the proportion of mature individuals in each age class ($Mat_{y,a}$) multiplied by the number of individuals in each age class:

@@

<!-- This method does not account for any changes in reproductive success by age, each kilogram of spawning biomass contributes the same to the total total reproductive output, irrespective of their age. Finally, the age-specific fecundity ($f_{y,a}$) is calculated as number of offspring produced by each individual in the stock.  -->

@@

$$
s_{y,a} =  s_{y,(a-1)} \times l_{y,(a-1)}
$$
The availability of catch-at-age time series in the data repository meant that fishing mortality and natural mortality rates were separable. Given this method provides separate estimates of $F$ and $M$ it was possible to calculate survival as well as theoretical population growth rates using the $M$ and fecundity ($\lambda_{F=0}$) estimates from the reconstruction. We recognize that the age structure in the underlying time series are still influenced by fishing pressure and that stocks are at varying size. This means that ($\lambda_{F=0}$) does not represent the maximum intrinsic rate of abundance increase [e.g., r_max; CITE]. However, our theoretical annual growth rates in the absence of fishing represent the productivity that the stock could have achieved in each specific year, if there had not been a fishery in that year.

@@

One limitation of calculating survivorship-to-age and fecundity-at-age using the archived data in the repository is that it is not straightforward to propagate observation error in each time series of data. To ensure that variability did not accumulate into unreasonable predictions for annual stock productivity, we fit the Euler-Lotka model using an iterative minimization process, where predicted abundance in the following year:

@@

Life table analyses make no assumptions about population regulation and thus do not incorporate a density-dependent relationship between spawners and recruits [@mcallisterUsingDemographicMethods2001]. However, our estimated productivity parameters could be explored for evidence of density-dependence by modelling each stock’s annual population growth rate against relative stock size (expressed as a proportion of maximum abundance) in the same year. Negative density dependence is observed when $\lambda$ increases as abundance declines [CITE] which is the type of compensatory response described by typical spawner-recruit relationships (ref, ref). Positive density dependence occurs when λ declines with increasing abundance, which is also called an Allee effect when it occurs at low abundance (ref, ref). Density independence occurs when λ shows no relationship with changes in abundance. We evaluated both the realized productivity estimates ($\lambda_{F>0}$) and the theoretical productivity values excluding the impact of fishing mortality ($\lambda_{F=0}$).

Life table analyses make no assumptions about population regulation and thus do not incorporate a density-dependent relationship between spawners and recruits (Reference?). However, the data could be explored for evidence of density-dependence by modelling each stock's estimated annual population growth rate against the relative stock size (proportion of maximum abundance) that year. This was done twice, once using  ($\lambda_{F=0}$) and once using ($\lambda_{F>0}$).  Negative density dependence is observed when $\lambda$ increases as abundance declines, positive density dependence occurs when $\lambda$ declines with increasing abundance, while density independence is observed when $\lambda$ is not impacted by changes in abundance.

@@

The annual fecundity and natural mortality values varied considerably for each stock. To account for this variability when evaluating recovery potential, we used a simulated stock projection. For `r n.dt.sims` simulations, we took a random sample of the annual values for fecundity and natural mortality, used the life table reconstruction to estimate $\lambda$, and calculated abundance in the following year. Abundance of each stock was projected over `r `n.sim.years` years, and the proportion of the simulations in which the stock size had doubled was recorded each year. We considered stocks to have high recovery potential if 75% of simulated trajectories doubled, medium potential if 50-74% of simulations doubled, and low recovery potential if < 50% of simulated trajectories doubled, calculated at 10 and 20 years into the projection. For simplicity, we did not incorporate density-dependent effects (e.g., potential population decline as abundance exceeds carrying capacity) or any correlation (either temporally or between fecundity and natural mortality). Once a population had doubled, it was considered to have remained at that abundance or greater for the remainder of the projection. This means that the metric gave the earliest possible doubling time for each simulation, assuming no future fishing mortality.

### 2024-12-17 Drafts/Paper_1/05-discussion.Rmd
Working on Discussion update
All comments from the summer are now integrated

Exploring demographic variability and realized compensatory responses in marine fish populations highlight several instances where stock dynamics do not conform to expected theoretical relationships, which has potentially serious implications for future management. Although most stocks exhibited a latent capacity for annual growth, and exhibited evidence for negative density dependence, overall theoretical growth rates in the absence of fishing were surprisingly close to 1, with several stocks having no apparent capacity for abundance increase in the majority of years. Quota-based fisheries management strategies often aim for neutral (50%) risk in the probability of population decline when setting harvest advice (ref, ref), so it was not unexpected that realized population growth rates centered around 1.

What was unsettling was how little capacity for increase these stocks seemed to retain when excluding the effects of fishing (i.e., $\lambda_{F=0}$). If fishing ceased, future recovery was likely within 20 years whenever the median  $\lambda_{F=0}$ was greater than 1.06 or the standard deviation of lambda was greater than 0.2. This applied to only ADD SOMETHING and none of the highest trophic level species. Widespread distributions, high egg production per female, and broadcast spawning strategies have all contributed to the impression that marine fishes should have very high potential for annual population growth [(CITE)]. When environmental variation results in favorable conditions, stocks are anticipated to exhibit high survival, resulting in large recruitment events or higher overall survival [(CITE)]. While we found indirect evidence of this, in that populations with high variability in  $\lambda_{F=0}$ also had high recovery potential, the magnitude of intra-species variability did not conform to expectations from conventional life-history theory [@kindsvaterTenPrinciplesEvolutionary2016]. For example, Atlantic herring are a small, relatively short lived, low trophic level pelagic species [CITE]. Their life history characteristics are considered to result in highly variable population growth potential which should allow them to exploit favorable environmental conditions and increase quickly in abundance [CITE]. However, the variability of  $\lambda_{F=0}$ for Atlantic herring was much lower than for the higher trophic level and longer lived Atlantic cod and haddock stocks in the same region. Furthermore, these three species co-occur in the NE Atlantic [CITE] and the timespan of data collection was similar, suggesting they likely experienced similar variability in oceanographic conditions. Other small pelagic populations, such as $A. tobianus$ and $T. esmarkii$ were characterized by higher  $\lambda_{F=0}$ values, which conforms to theoretical expectations for small, low trophic level species. However, these two stocks were found in the North Sea, which appears to be an environment in which high variability in  $\lambda_{F=0}$ is commonplace for harvested stocks. In addition, while stocks with a high variability in $\lambda_{F=0}$ do have years with very high population growth, these events happen at irregular intervals, even when populations are at low abundances. Contrary to assumptions from typical stock-recruit relationships [CITE], stocks do not necessarily exhibit increased compensatory capacity when at low abundance. These data suggest that while most stocks have the capacity for rapid abundance increase, they only do so occasionally even when at low abundance. Thus, accounting for both the magnitude and frequency of years with high population growth become key components during the development of fisheries advice, as well as for the development of recovery plans and conservation targets [CITE ref new Fisheries act stuff? Precautionary approach stuff?].

NOTE THAT THE BECAUSE WE ASSUMED FISHERY REMOVALS WERE CORRECT COMBINED WITH THE PRODUCITIVTY PARAMETERS HAVING RESULTED IN LAMBDA VALUES THAT ARE 5% LOWER, IF THAT DECLINE IN LAMBDA WAS DUE TO FISHING, OUR PROJECTIONS, GENERATION TIME, AND R0 VALUES WOULD ALL BE SOMEWHAT CONSERVATIVE AS TO GET THAT LOWER LAMBDA VALUE WE ATTRIBUTED THIS TO THE PRODUCTIVITY OF THE SYSTEM BEING LOWER, WHEREAS ATTRIBUTING THAT DIFFERENCE SOLELY TO FISHING WOULD HAVE HAD NO IMPACT ON THE NATURAL MORTALITY OR FECUNDITY

Our best understanding of productivity in these assessed fish stocks suggests that equilibrium theory and general life history characteristics (e.g. longevity, trophic level) are poor predictors of population growth rates and recovery potential. For stocks that have collapsed or are considered to be at very low abundance, it would be prudent to evaluate lifetime reproductive rates prior to developing harvest advice. Numerous stocks in our analysis had no apparent capacity to increase in abundance even in the absence of fishing, which would assure further declines if harvesting continued. For example, there were seven populations in which $\lambda_{F=0}$ at 40% of maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* [3] and *G. chalcogrammus* [2]). These stocks exhibited characteristics much more similar to endangered populations [e.g., several Atlantic salmon Designatable Units in Canada, CITE COSEWIC REPORT], which can also have lifetime reproductive rates < 1 [Gibson refs, CITE], in addition, there was little evidence of negative density dependence for these stocks. In our analysis, the most marked example was Atlantic cod in the NW Atlantic. We have little understanding of how large these Atlantic cod stocks would have been originally. The maximum abundance and thus the magnitude of decline used for our analyses represents the outcome of > 100 years of industrial fishing [Cite Rose book maybe?, others]. Similar to salmon, extreme abundance decline combined with changes in the environment [Cite cold water and seals] may have resulted in these stocks having little latent capacity for growth or recovery. Indeed, it has been suggested that the SGSL cod population will go extinct unless natural mortality declines significantly [CITE Doug’s predator pit stuff], a conclusion supported by our analysis. However, the low recovery potential of Atlantic cod was no longer observed as one moves towards the eastern Atlantic, with stocks around mainland Europe all exhibiting high recovery potential. As with $G. chalcogrammus$, this indicates that regional variability must factor strongly into a species’ ability to withstand fishing pressure. Given that variability in population growth potential among stocks could be substantially greater than among species (e.g., compare the variability of $\lambda$ between Atlantic cod and Atlantic herring), life history characteristics and/or trophic level appear to be poor predictors of how marine teleosts will respond to fishing. This problem would compound for less data-rich stocks, in that typical assumptions made for data-poor assessments (give examples of assessment methods) would be more likely to lead to inappropriate harvest advice. That life history characteristics did not link strongly to expected productivity is extremely concerning, given the need for sustainable fisheries management of data-poor stocks [CITE] to mitigate risks of overfishing across the globe [CITE].

Our life-table reconstruction enables the underlying demographic rates to be estimated for populations with sufficient age-structured data, this provides an opportunity of understanding the dynamics of these populations using concepts developed in conservation biology and ecology. In doing so, our optimization procedure assumed that both the estimates of abundance and the fishery removals from the stock assessment are correct. Although these assumptions are unlikely to be fully met [CITE], it is important to remember that harvest advice is being derived for these stocks under the same general assumptions. Thus, our analyses reveal that current assessments can be based on vastly different understandings of productivity and the potential for population growth for the same species, even when incorporating similar elements to describe population dynamics (e.g., equilibrium theory). Our analysis suggests that management advice might be fairly accurate for stocks that aren’t overly variable or those that retain high latent capacity for population increase. Low variability would suggest that the stock is not overly influenced by variation in environmental conditions [or that there was little environmental variability in the ecosystem, CITE], while high latent capacity for population increase could indicate the opposite [CITE]. However, it is not immediately apparent what conditions (environmental, ecological and/or anthropogenic) might be the primary determinants of population growth rates for a given cohort or system. Our findings underscore the need for alternative frameworks in which to derive advice, particularly those that incorporate consideration of interspecies relationships and/or environmental conditions under the umbrella of ecosystem-based fishery management [CITE]. They also underscore the need to collect stock-specific data on marine teleosts, rather than using species-level information from global repositories [e.g., Fishbase, Fishlife, CITE] or assuming values from closely related species to derive harvest advice [CITE].

@@

Here we deconstructed the ecological demographics of numerous harvested marine fish populations . In general, these populations have latent capacity for growth and, even when accounting for the influence of fishing, these growth rates are very close to 1 in general.

For every population in this analysis the $\lambda_{F=0}$ at the minimum observed abundance was estimated to be greater than one; suggesting all of the populations have the potential for some degree of recovery given enough time when there is no fishing mortality. However, there are seven populations in which $\lambda_{F=0}$ at 40% of the maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* (3) and *G. chalcogrammus* (2)). For both of these Gadus species, there are examples of populations in which the latent capacity for growth is far more robust.  For example, *G. chalcogrammus* is one of the largest fisheries in the world and the vast majority of the landings are from the Eastern Bering Sea population which has a high latent capacity for growth. However the $\lambda_{F=0}$ estimates for two neighboring populations in Alaska are much lower, and it is likely that these populations experience much lower fishing pressure because of their low latent capacity for growth.

>>>>

Exploring demographic variability and realized compensatory responses in marine fish populations highlight several instances where stock dynamics do not conform to expected theoretical relationships, which has potentially serious implications for future management. Although most stocks exhibited a latent capacity for annual growth, and exhibited evidence for negative density dependence, overall theoretical growth rates in the absence of fishing were surprisingly close to 1, with several stocks having no apparent capacity for abundance increase in the majority of years. Quota-based fisheries management strategies often implicitly or explicitly aim for neutral (50%) risk in the probability of population decline when setting harvest advice [e.g., @marentetteCrossJurisdictionalReviewInternational2020; @dfoScienceAdvicePrecautionary2021], so it was not unexpected that realized population growth rates centered around 1 when the impact of fishing activity was included.

What was unsettling was how little capacity for increase these stocks seemed to retain when excluding the effects of fishing (i.e., $\lambda_{F=0}$). If fishing ceased, future recovery was likely within 20 years whenever the median  $\lambda_{F=0}$ was $\gtrsim$ lam.rebuild.safe or the standard deviation of lambda was $\gtrsim$ `r sd.rebuild.safe`.  Widespread distributions, high egg production per female, and broadcast spawning strategies have all contributed to the impression that marine fishes should have very high potential for annual population growth [CITE]. When environmental variation results in favorable conditions, stocks are anticipated to exhibit high survival, resulting in large recruitment events or higher overall survival [CITE SOME BET HEDGING IN FISH STUFF?]. While we found indirect evidence of this, in that populations with high variability in  $\lambda_{F=0}$ also had high recovery potential, the magnitude of intra-species variability did not conform to expectations from conventional life-history theory [@kindsvaterTenPrinciplesEvolutionary2016]. For example, Atlantic herring are a small, relatively short lived, low trophic level pelagic species [CITE]. Their life history characteristics are considered to result in highly variable population growth potential which should allow them to exploit favorable environmental conditions and increase quickly in abundance [CITE]. However, the variability of  $\lambda_{F=0}$ for Atlantic herring was much lower than for the higher trophic level and longer lived Atlantic cod and haddock stocks in the same region. Furthermore, these three species co-occur in the NE Atlantic [CITE] and the timespan of data collection was similar, suggesting they likely experienced similar variability in oceanographic conditions. Other small pelagic populations, such as $A. tobianus$ and $T. esmarkii$ were characterized by higher  $\lambda_{F=0}$ values, which conforms to theoretical expectations for small, low trophic level species. However, these two stocks were found in the North Sea, which appears to be an environment in which high variability in  $\lambda_{F=0}$ is commonplace for harvested stocks. In addition, while stocks with a high variability in $\lambda_{F=0}$ do have years with very high population growth, these events happen at irregular intervals, even when populations are at low abundances. Contrary to assumptions from typical stock-recruit relationships [CITE], stocks do not necessarily exhibit increased compensatory capacity when at low abundance. These data suggest that while most stocks have the capacity for rapid abundance increase, they only do so occasionally even when at low abundance. Thus, accounting for both the magnitude and frequency of years with high population growth become key components during the development of fisheries advice, as well as for the development of recovery plans and conservation targets [@marentetteCrossJurisdictionalReviewInternational2020; @dfoScienceAdvicePrecautionary2021].

Our best understanding of productivity in these assessed fish stocks suggests that equilibrium theory and general life history characteristics (e.g. longevity, trophic level) are poor predictors of population growth rates and recovery potential. For stocks that have collapsed or are considered to be at very low abundance, it would be prudent to evaluate lifetime reproductive rates prior to developing harvest advice. Numerous stocks in our analysis had no apparent capacity to increase in abundance even in the absence of fishing, which would assure further declines if harvesting continued. For example, there were seven stocks in which $\lambda_{F=0}$ at 40% of maximum abundance is below 1, five of these stocks are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* [3] and *G. chalcogrammus* [2]). These stocks exhibited characteristics much more similar to endangered stocks [e.g., several Atlantic salmon Designatable Units in Canada, CITE COSEWIC REPORT], which can also have lifetime reproductive rates < 1 [@ gibsonEquilibriumAnalysesPopulation2009; @bowlbyReductionFitnessLimits2011, CITE], in addition, there was little evidence of negative density dependence for these stocks. In our analysis, the most marked example was Atlantic cod in the NW Atlantic. We have little understanding of how large these Atlantic cod stocks would have been originally. The maximum abundance and thus the magnitude of decline used for our analyses represents the outcome of > 100 years of industrial fishing [Cite Rose book maybe?, others]. Similar to salmon, extreme abundance decline combined with changes in the environment [Cite cold water and seals] may have resulted in these stocks having little latent capacity for growth or recovery. Indeed, it has been suggested that the SGSL cod stock will go extinct unless natural mortality declines significantly [CITE Doug’s predator pit stuff], a conclusion supported by our analysis. However, the low recovery potential of Atlantic cod was no longer observed as one moves towards the eastern Atlantic, with stocks around mainland Europe all exhibiting high recovery potential. As with $G. chalcogrammus$, this indicates that regional variability must factor strongly into a species’ ability to withstand fishing pressure. Given that variability in stock growth potential among stocks could be substantially greater than among species (e.g., compare the variability of $\lambda$ between Atlantic cod and Atlantic herring), life history characteristics and/or trophic level appear to be poor predictors of how marine teleosts will respond to fishing. This problem would compound for less data-rich stocks, in that typical assumptions made for data-poor assessments (give examples of assessment methods) would be more likely to lead to inappropriate harvest advice. That life history characteristics did not link strongly to expected productivity is extremely concerning, given the need for sustainable fisheries management of data-poor stocks [CITE] to mitigate risks of overfishing across the globe [CITE].

Our life-table reconstruction enables the underlying demographic rates to be estimated for stocks with sufficient age-structured data, this provides an opportunity of understanding the dynamics of these stocks using concepts developed in conservation biology and ecology. In doing so, our optimization procedure assumed that both the estimates of abundance and the fishery removals from the stock assessment are correct. Although these assumptions are unlikely to be fully met [CITE], it is important to remember that harvest advice is being derived for these stocks under the same general assumptions. Thus, our analyses reveal that current assessments can be based on vastly different understandings of productivity and the potential for population growth for the same species, even when incorporating similar elements to describe population dynamics (e.g., equilibrium theory). Our analysis suggests that management advice might be fairly accurate for stocks that aren’t overly variable or those that retain high latent capacity for population increase. Low variability would suggest that the stock is not overly influenced by variation in environmental conditions [or that there was little environmental variability in the ecosystem, CITE], while high latent capacity for population increase could indicate the opposite [CITE]. However, it is not immediately apparent what conditions (environmental, ecological and/or anthropogenic) might be the primary determinants of population growth rates for a given cohort or system. Our findings underscore the need for alternative frameworks in which to derive advice, particularly those that incorporate consideration of interspecies relationships and/or environmental conditions under the umbrella of ecosystem-based fishery management [CITE]. They also underscore the need to collect stock-specific data on marine teleosts, rather than using species-level information from global repositories [e.g., Fishbase, Fishlife, CITE] or assuming values from closely related species to derive harvest advice [CITE].

There are some limitations to this kind of life table reconstruction. Given we use historic data sets, the nature of the population growth are somewhat constrained, for example, when the stock is at the highest abundance in the observed time series, $\lambda$, by definition, has to be $\le$ 1, conversely at the lowest abundance observed $\lambda$ must be $\ge$ 1. There are more subtle constraints on the lambda estimates, for example, $\lambda$ must be $\le$ 2 when a stock is at 50% of the maximum abundance, and the minimum value of $\lambda$ similarly constrained by the minimun observed abundance observed in the time series. These constraints contributed to an increase in the variability of $lambda$ observed in the analysis.  In addition we assumed that the fishery removals from the stock assessments were correct and modified the productivity parameters to match the time series from the stock assessment; this resulted in $\lambda$ values that were 5% lower than expected using the initial productivity parameters from the stock assessment. If this decline in $\lambda$ was instead attributed to fishing mortality, then the underlying productivity of these systems would have resulted in less conservative projections (DOUBLING TIME... KEEP PULLING ON THIS THEAD
)
NOTE THAT THE BECAUSE WE ASSUMED FISHERY REMOVALS WERE CORRECT COMBINED WITH THE PRODUCITIVTY PARAMETERS HAVING RESULTED IN LAMBDA VALUES THAT ARE 5% LOWER, IF THAT DECLINE IN LAMBDA WAS DUE TO FISHING, OUR PROJECTIONS, GENERATION TIME, AND R0 VALUES WOULD ALL BE SOMEWHAT CONSERVATIVE AS TO GET THAT LOWER LAMBDA VALUE WE ATTRIBUTED THIS TO THE PRODUCTIVITY OF THE SYSTEM BEING LOWER, WHEREAS ATTRIBUTING THAT DIFFERENCE SOLELY TO FISHING WOULD HAVE HAD NO IMPACT ON THE NATURAL MORTALITY OR FECUNDITY

@@

Here we deconstructed the ecological demographics of numerous harvested marine fishs . In general, theses have latent capacity for growth and, even when accounting for the influence of fishing, these growth rates are very close to 1 in general.

For every in this analysis the $\lambda_{F=0}$ at the minimum observed abundance was estimated to be greater than one; suggesting all of thes have the potential for some degree of recovery given enough time when there is no fishing mortality. However, there are seven populations in which $\lambda_{F=0}$ at 40% of the maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* (3) and *G. chalcogrammus* (2)). For both of these Gadus species, there are examples of populations in which the latent capacity for growth is far more robust.  For example, *G. chalcogrammus* is one of the largest fisheries in the world and the vast majority of the landings are from the Eastern Bering Sea population which has a high latent capacity for growth. However the $\lambda_{F=0}$ estimates for two neighboring populations in Alaska are much lower, and it is likely that these populations experience much lower fishing pressure because of their low latent capacity for growth.

### 2024-12-17 Drafts/Paper_1/index.Rmd
Working on Discussion update
All comments from the summer are now integrated

---
title: |
   "Life history variability influence on the recovery potential of globally exploited teleosts "
french_title: "Titre ici (*Nom latin de l'espèce*)"
author: |
  David. M. Keith^1^ 
  Freya M. Keyser^1^
  Julie M. Charbonneau^2^
  Heather 'The Boss' Bowlby^1^
author_list: "Keith, D.M., Keyser, F.M., Charbonneau, J.M., Bowlby, H.T.B."
address: |
  ^1^Bedford Institute of Oceanography\
     Fisheries and Oceans Canada, 1 Challenger Dr.\
     Dartmouth, Nova Scotia, B2Y 4A2 , Canada\
  \smallskip
  ^2^Simon Fraser University\
     Department of Biology\
french_address: |

@@

abstract: |
  Here is the abstract text. Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

  Start new paragraphs after a blank line and with 2 spaces indent. Lorem ipsum dolor sit amet, consectetur adipisicing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.

>>>>

---
title: |
   "Variability in the recovery potential of globally exploited marine teleosts "
french_title: "Titre ici (*Nom latin de l'espèce*)"
author: |
  David. M. Keith^1^ 
  Freya M. Keyser^1^
  Julie M. Charbonneau^2^
  Heather 'The Boss' Bowlby^1^
author_list: "Keith, D.M., Keyser, F.M., Charbonneau, J.M., Bowlby, H.T.B."
address: |
  ^1^Bedford Institute of Oceanography\
     Fisheries and Oceans Canada, 1 Challenger Dr.\
     Dartmouth, Nova Scotia, B2Y 4A2 , Canada\
  \smallskip
  ^2^Simon Fraser University\
     Department of Biology\
french_address: |

@@

abstract: |
  Marine fishes were once believed to be extremely resilient to the effects of fishing. However, this view has changed over time, partly due to the collapse and subsequent lack of recovery of several histically important stocks across the globe. This has raised questions about the underlying assumptions used in the management of these stocks and the impact of environmental factors on the population dynamics. Despite extensive research exploring these challenges, few studies have explored the underlying demographic variability of harvested marine fishes using conservation biology and population ecology perspectives. Here we reconstruted the demographics of 77 stocks that are assessed in the north Atlantic and the northeast Pacific Oceans using life-tables. This method allowed for the exploration of the demographic variability of these stocks using metrics commonly used in conservation biology (i.e., population growth rates, doubling time, lifetime reproductive success, and generation time).In general, these stocks had a latent capacity for population growth (median $\lambda$ = 1.1) when the effect of fishing was excluded. When the impacts of fishing were accounted for the stock mean $\lambda$ was 1.02 overall. However, $\lambda$ when accounting for fishing were less than one 57% of the time. As a result, for many of the stocks in this analysis, population growth is driven by years with exceptionally high population growth rates.  For most stocks, as abundance declined, $lambda$ tended to increase and become more variable. These general patterns are highly variable among stocks and regionally, with notable differences within Atlantic cod stocks in the northwest (NW) Atlantic. These stocks had few large recruitment events and evidence of positive density dependence (i.e., Allee effects or depensation) was relatively strong. The recovery potential was relatively high, with simulations indicating that 60 of the stocks would be highly likely to double in size within 20 years in the absence of fishing. Stocks with a low recovery potential were constrained to stocks with both a low median $\lambda$ and low variability in the $\lambda$. While most stocks exhibited a latent population growth capacity in the absence of fishing, regional and stock specific factors revealed that several stocks lacked this capacity, even in the absence of fishing. Knowledge of the underlying demographic parameters for a stock can be used to help sustainably manage the stock and can inform the development of rebuilding plans for stocks that have collapsed.  Finally, we note that life history characteristics alone were a poor predictor of annual population growth or future recovery potential, with greater variability observed among stocks rather than among species with distinct regions associated with generally higher or generally lower vital rates.
